---
title: "Estimating Movement Paths and Haul-out Behavior of Harbor Seals (_Phoca vitulina_) in Cook Inlet, Alaska"
format: 
  html:
    theme: cosmo
    mainfont: 'Public Sans'
toc: true
toc-depth: 2
crossref:
  tbl-prefix: Table
  fig-prefix: Figure
editor: visual
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Public+Sans:wght@300&display=swap" rel="stylesheet">

::: callout
## Josh M. London

*Wildlife Biologist*, Marine Mammal Laboratory\
Alaska Fisheries Science Center, NOAA Fisheries\
Seattle, Washington
:::

::: {.callout-note icon="false"}
## Summary

This document provides a description of the process used to import and tidy
telemetry data observations from bio-loggers deployed on harbor seals in Cook
Inlet, Alaska. The data include estimated location observations and percent-dry
timelines that indicate the percentage of each hour the bio-logger was dry. The
location observations are used to model the movement parameters and predict a
likely track for each seal. Each of these movement paths is examined for regions
that cross-land (not sensible for a harbor seal) and those sections are
re-routed around the land. The initial movement model is developed using the R
package `crawl` and the paths are re-routed around land with the R package
`pathroutr`. The final data product is an `sf` spatial data frame of predicted
points every hour and, when available, the associated percent-dry percentage for
that hour --- this indicates what percentage of the hour the seal was likely
hauled out on land resting.
:::

```{r setup, include=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  include = FALSE,
  dev = "ragg_png"
)

library(targets)
library(dplyr)
library(tidyr)
library(sf)
library(ggspatial)
library(gt)
library(ggplot2)
library(MetBrewer)

options(scipen = 3, digits = 3)

theme_set(theme_minimal(base_family = 'Public Sans'))
theme_replace(plot.title.position = "plot")
```

## Importing & Tidy-ing the Telemetry Data

### Background

```{r summary-tbl-data, include=TRUE}

summary_tbl <- tar_read(locs_data) %>% 
  as_tibble() %>% 
  select(deployid, speno, sex, age, tag_family, deploy_dt, end_dt) %>% 
  group_by(speno, sex, age) %>% 
  summarise(n_tags = n_distinct(tag_family), 
            deploy_start = min(deploy_dt),
            deploy_end = max(end_dt)
            ) %>% 
  group_by(sex, age) %>% 
  summarise(n_tags = sum(n_tags),
            n_seals = n())
```

Between 2004 and 2006, `r sum(summary_tbl$n_seals)` were captured and released
with satellite-linked bio-loggers attached as part of a research effort in Cook
Inlet, Alaska. Each transmitter was glued to the hair on the back of the seal
using durable epoxy. Fourteen of the seals were also equipped with specially
developed transmitters that were attached to one of the rear flippers. 
@tbl-summary provides more information on the distriubtion of deployments
across sex and age classes.
Transmissions from the `r sum(summary_tbl$n_tags)` bio-loggers resulted in 
`r nrow(tar_read(locs_data))` location estimates and 
`r nrow(tar_read(timeline_data))` haul-out behavior records. Harbor seals captured
in central and southern Cook Inlet ranged as far southwest as the Semidi Islands
and some seals used habitats around the north and northeast side of Kodiak
Island. There was a strong seasonal pattern of more coastal and restricted
spatial use during the spring and summer (breeding, pupping, molting) and more
wide-ranging movements within and outside of Cook Inlet during the winter
months. These insights into seasonal harbor seal movements, behavior and
abundance are critical for understanding foraging ecology, marine habitat use,
and risk of impacts from industrial accidents such as oil spills.

```{r tbl-summary, echo=FALSE, include=TRUE}
#| tbl-cap: "The number of bio-loggers deployed on harbor seals in
#| Cook Inlet, Alaska between 2004 and 2006 across sex and age class.
#| A few seals were released with two devices attached; one adhered to
#| the hair on dorsal side and another attached through the rear flipper
#| inter-digital webbing."

summary_tbl %>% 
  mutate(sex = case_when(
    sex == "F" ~ "Female",
    sex == "M" ~ "Male"
  )) %>% 
  gt(
    rowname_col = "age",
    groupname_col = "sex"
    ) %>% 
  tab_header(
    title = "Bio-logger Deployments on Harbor Seals") %>% 
  tab_footnote(
    footnote = "some seals were released with two bio-logger devices",
    locations = cells_column_labels(
      columns = n_tags
    )
  ) %>% 
  tab_footnote(
    footnote = "age class 'YEARLING' is often combined with 'SUBADULT",
    locations = cells_stub(rows = "YEARLING")
  ) %>% 
  summary_rows(
    groups = TRUE,
    columns = c(n_tags, n_seals),
    fns = list(total = "sum"),
    missing_text = "",
    formatter = fmt_integer
  ) %>% 
  grand_summary_rows(
    columns = c(n_tags, n_seals),
    fns = list(total = "sum"),
    formatter = fmt_integer,
    missing_text = ""
  ) %>% 
  tab_spanner(
    label = "Sample Size",
    columns = c(`n_tags`, `n_seals`)
  ) %>%
  cols_label(
    sex = "Sex",
    age = "Age Class",
    n_tags = "# of tags",
    n_seals = "# of seals"
  ) %>%
  tab_style(
    style = list(
      cell_text(
        weight = "bold"
      ),
      cell_borders(
        sides = "bottom",
        color = "darkgrey",
        weight = px(2)
      ),
      cell_borders(
        sides = "top",
        weight = px(0)
      )
    ),
    locations = list(
      cells_row_groups()
    )
  ) %>% 
  tab_style(
    style = list(
      cell_text(
        weight = "bold"
      )
    ),
    locations = list(
      cells_grand_summary(),
      cells_stub_grand_summary()
    )
  ) %>% 
  tab_style(
    style = list(
      cell_borders(
        sides = "top",
        color = "darkgrey",
        weight = px(2)
      )
    ),
    locations = list(
      cells_grand_summary(),
      cells_stub_grand_summary()
    )
  ) %>% 
  tab_style(
    style = list(
      cell_borders(
        sides = "bottom",
        weight = px(0)
      )
    ),
    locations = list(
      cells_summary()
    )
  ) %>% 
  tab_options(
    table.width = pct(90),
    table_body.hlines.width = 0,
    stub.border.width = 0,
    summary_row.background.color = "grey90",
    table.font.size = "medium",
    data_row.padding = px(3),
    summary_row.padding = px(3),
    grand_summary_row.padding = px(3)
  )
```

### Data Handling and Management

All the processes and code used for this project are [available as a repository
on GitHub](https://github.com/jmlondon/cookinlet-seal-movt). Data retrieval,
processing, analysis are all managed as a _pipeline_ via the `targets` package in
R. For a more detailed exploration of the code and process, examine the 
[_targets.R](https://github.com/jmlondon/cookinlet-seal-movt/blob/main/_targets.R) file
and the functions [within the project's R folder](https://github.com/jmlondon/cookinlet-seal-movt/tree/main/R).


All of the tags deployed on seals in Cook Inlet were manufactured by
[Wildlife Computer](https://wildlifecomputers.com) and originally processed
through their data portal. To facilitate access to data from the portal
within R, we relied on the `wcUtils` R package which 
[is available from GitHub](https://github.com/jmlondon/wcUtils). All telemetry
data within the Polar Ecosystems Program (PEP) is maintained within a PostgreSQL
database. This also includes important details regarding the individual seal
(e.g. _speno_, age class, sex, morphometrics) and additional deployment metadata for
each bio-logger (e.g. _deployid_, deployment start/end times, tag serial number).
Initial data were queried from the database and, for the location observations,
a course speed filter was applied (7.5 meters/sec) to remove any egregiously
wrong location estimates. The percent-dry timeline data were mostly as is from
the database

::: {.callout-caution}
The current version of the data delivered via `pins` **does not group
deployments by _speno_**. Instead each deployment is treated as a separate
individual seal. This is problematic for those seals that were released
with two bio-logging devices. This will be resolved in the next major iteration
of the data published via `pins`.
:::

### Mapping the Observed Location Data

First, we will take a look at the spatial distribution of the observed location
estimates (this is after the course speed filter has been applied). In
@fig-obs-map, the observed locations are concentrated within the Cook Inlet
region --- as would be expected --- but, note, there is considerable expansion
of the range into the Shelikof Strait, Kodiak, and out toward the shelf break in
the Gulf of Alaska. The error associated with observed locations is also evident
as the spatial distribution extends well inland beyond what would be sensible
for harbor seals.

```{r obs-hexbins}
map_data <- tar_read(locs_data) %>% 
  dplyr::mutate(year = lubridate::year(locs_dt)) %>% 
  dplyr::filter(year %in% c("2004", "2005", "2006")) %>% 
  sf::st_crop(sf::st_bbox(tar_read(land_osm)))

hexgrid <- sf::st_make_grid(st_bbox(map_data) %>% st_as_sfc(), cellsize = 25*1000,
                            what = "polygons", square = FALSE) 
hexgrid <- st_sf(index = 1:length(lengths(hexgrid)), hexgrid)

hexbin <- st_join(map_data, hexgrid, join = st_intersects)

locs_count <- hexgrid %>%
  left_join(
    count(hexbin, index) %>%
      as_tibble() %>%
      dplyr::select(index, ct=n)
  ) %>% tidyr::drop_na()
```

```{r, include=TRUE, echo = FALSE, fig.height=6}
#| label: fig-obs-map
#| fig-cap: "Spatial distribution of observed location estimates from
#| bio-loggers deployed on harbor seals in Cook Inlet from 2004-2006."

esri_ocean <- paste0('https://services.arcgisonline.com/arcgis/rest/services/',
                     'Ocean/World_Ocean_Base/MapServer/tile/${z}/${y}/${x}.jpeg')

ggplot() +
  annotation_map_tile(type = esri_ocean, zoomin=0, progress = "none") +
  layer_spatial(
    data = locs_count, size = 0.125,
    aes(fill = ct, alpha = 0.35)
  ) +
  scale_fill_gradientn(colors = rev(met.brewer("Greek")),
                    trans = "log10", aesthetics = "fill",
    name = "Number of locations",
    guide = guide_colorbar(barwidth = 1,title.position = "right",
                           title.hjust = 0.5,
                           barheight = 12)
    ) + 
  theme(legend.title = element_text(angle = -90)) +
  guides(alpha="none") +
  ggtitle("Spatial distribution of observed location data") +
  labs(caption = bquote("One hexagonal cell represents 25"~km^2))
```


### Distribution of Data Across Time

As with spatial distribution, it can also be informative to explore how the
amount of data varied over time. This is especially important with bio-logging
data because not all deployments start at the same time and deployment end
dates are determined by a combination of factors (e.g. battery life, tag
failures, adhesive failures, molt, and animal behavior). For this research
project, there were two research cruises per year. One was timed to occur just
after molt (September-October) so devices adhered to the hair would have potential to last for
9-12 months. The other cruise was timed for May to correspond with the 
expected median tag duration from the fall cruise. This also ensured a good
sample size of deployments would be active during the pupping season (~June).


```{r locs-day-tbl, include=TRUE}
locs_day_tbl <- tar_read(locs_data) %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(yday = lubridate::yday(locs_dt),
                year = lubridate::year(locs_dt)) %>% 
  dplyr::filter(year %in% c("2004", "2005", "2006")) %>% 
  dplyr::group_by(yday) %>% 
  dplyr::count()
```


```{r, include=TRUE, echo=FALSE}
#| label: fig-locs-by-day
#| fig.height: 3
#| fig-cap: "Distribution of observed location data by day-of-year and
#| pooled across all years (2004-2006). The gap in August corresponds to the annual
#| molt when seals grow a new coat of hair and bio-loggers adhered to the hair
#| are shed. The higher intensity of location data in the spring and fall
#| is reflective of the timing of capture cruises (typically May and September)."

ggplot(data = locs_day_tbl, aes(x = yday, y = 1, fill = n))+
        geom_tile()+
        scale_fill_gradientn(colors = rev(met.brewer("Hokusai2"))) +
  scale_x_continuous(breaks = c(1, 31, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
                           labels = c("01 Jan","01 Feb", "01 Mar", "01 Apr", "01 May", 
                                      "01 Jun", "01 Jul", "01 Aug","01 Sep",
                                      "01 Oct", "01 Nov", "01 Dec"),
                           expand = expansion(mult = c(0, 0))) +
        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  guides(fill = guide_colorbar(title.position = "bottom", 
                               title.hjust = 0.5, barwidth = 15,
                               barheight = 0.5)) +
  theme(axis.text.x = element_text(hjust = 0),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(fill = "number of locations per day") +
  ggtitle("Density of location observations varies across the calendar year",
          subtitle = "data are presented as the number of locations per day collated across years (2004-2006)")
```

The density of locations presented in @fig-locs-by-day are pooled across all
the years of the study. But, within a given year, there are larger gaps in
data coverage that are important to note. @fig-locs-by-day-year presents the
density of location data by year and provides a more insightful view. For
instance, most of the data in the study for the months of January through
April come from deployments in 2006. Only the months of May-July and 
September-December are represented in more than one year --- all of which is to
be expected given the original objectives of the study, typical duration of
bio-loggers, and the timing of research cruises.

```{r locs-day-yr-tbl, include=TRUE}
locs_day_year_tbl <- tar_read(locs_data) %>% 
  tibble::as_tibble() %>% 
  dplyr::mutate(yday = lubridate::yday(locs_dt),
                year = lubridate::year(locs_dt)) %>% 
  dplyr::filter(year %in% c("2004", "2005", "2006")) %>% 
  dplyr::group_by(year, yday) %>% 
  dplyr::count()
```

```{r, include=TRUE, echo=FALSE}
#| label: fig-locs-by-day-year
#| fig.height: 5
#| fig-cap: "Distribution of observed location data by day-of-year for
#| each of the years in the study (2004-2006). The gaps (white space) correspond
#| to periods of time within the year when no bio-loggers were transmitting. 
#| The higher intensity of location data in the spring and fall
#| is reflective of the timing of capture cruises (typically May and September)."

ggplot(data = locs_day_year_tbl, aes(x = yday, y = 1, fill = n))+
        geom_tile()+
        scale_fill_gradientn(colors = rev(met.brewer("Hokusai2"))) +
  scale_x_continuous(breaks = c(1, 31, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
                           labels = c("01 Jan","01 Feb", "01 Mar", "01 Apr", "01 May", 
                                      "01 Jun", "01 Jul", "01 Aug","01 Sep",
                                      "01 Oct", "01 Nov", "01 Dec"),
                           expand = expansion(mult = c(0, 0))) +
        scale_y_continuous(expand = expansion(mult = c(0, 0))) +
  guides(fill = guide_colorbar(title.position = "bottom", 
                               title.hjust = 0.5, barwidth = 15,
                               barheight = 0.5)) +
  facet_grid(year ~ .) +
  theme(axis.text.x = element_text(hjust = 0),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom") +
  labs(fill = "number of locations per day") +
  ggtitle("Density of location observations varies across the years",
          subtitle = "data are presented as the number of locations per day")
```

### Percent-dry Timeline Data

In addition to the location estimates, the bio-loggers in this study also
recorded the percentage of each hour the wet/dry sensor was dry. This is a way
to explore haul-out behavior because, other than a few rare instances when a
seal rests on the surface for extended periods, the percent-dry can be
interpreted as percent of the hour hauled out. Haul-out behavior data were
recorded and transmitted via the Argos satellite network as hourly percent-dry
timelines. For each hour of a day, the wet/dry sensor was polled by the tag
firmware every few seconds and the percent of the hour in the dry state was
calculated. On board the bio-logger, hourly percent-dry data were rounded to the
nearest 10% inclusive of 0% and 100% and additional values at 3% and 98%.

```{r, include = TRUE, echo = FALSE}
#| label: fig-percent-dry
#| fig-cap: "Haul-out behavior observations recorded by a bio-logger deployed
#| on a harbor seal over two years. Areas of the plot with white space represent
#| missing data that were not successfully received from the tag."

d <- tar_read(timeline_data) %>% filter(deployid == "PV2005_0267_05L0083") %>% 
  dplyr::select(speno, timeline_start_dt, percent_dry) %>% 
  arrange(timeline_start_dt) %>% 
  mutate(year = lubridate::year(timeline_start_dt),
         month = lubridate::month(timeline_start_dt,label=TRUE),
         day = lubridate::day(timeline_start_dt),
         hour = lubridate::hour(timeline_start_dt))

p <- ggplot(d,aes(day,hour,fill=percent_dry))+
  geom_tile(color= "white",size=0) + 
  scale_fill_gradientn(colors = rev(met.brewer("Hiroshige")),
                       name = "hourly percent dry", aesthetics = "fill",
    guide = guide_colorbar(title.position = "bottom", barwidth = 15,
                           barheight = 0.5, title.hjust = 0.5)
    )

p <- p + facet_grid(year~month, scales = "free_x")
p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
  scale_y_continuous(breaks = c(4,12,20))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  theme(panel.spacing = unit(0,"line")) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry actogram for a single harbor seal across two years") +
  xlab("day of month") + ylab("hour (UTC)")
p
```

::: {.callout-note}
There's currently an issue with the figure above b/c there's missing data
for entire months of July and August. Will need to supplement the data
record with missing values to make the plot consistent.
:::

## Movement Modeling & Predicting Paths

### Fit movement models with `crawl` package

```{r, file = 'R/fit_crawl.R', include=TRUE, eval=FALSE, echo=TRUE}

```

### Re-fit any problematic deployments

```{r, file = 'R/refit_crawl.R', include=TRUE, eval=FALSE, echo=TRUE}

```


## Re-routing Paths Around Land Barriers
